---
title: "Task 3 Sample analysis"
author: "Will Wright"
format: 
  html:
    code-fold: false
    theme: cosmos
---

In this notebook, I run through the climate change analysis for abundance, using the model made in Task 2.

# Helper functions

```{r}
#| message: false
#| warning: false
#| label: "helper_functions"
#| code-fold: true
get_climate_predictions <- function(
  wetland,
  model,
  spp_code = NULL,
  fx_group = NULL,
  flow_metadata = combined_flow_metadata,
  climate_projections = valley_climate_projections_with_wetlands,
  basin_flow_stats = basin_flow_mean_sd
) {
  # --- Pre-processing -------------------------------------------------------
  scaling <- combined_flow_metadata |>
    filter(wetlands == wetland) |>
    slice(1) |>
    select(
      wetland_mean_flow_3m,
      wetland_sd_flow_3m,
      wetland_mean_flow_annual,
      wetland_sd_flow_annual
    )

  # --- Build scenario newdata -----------------------------------------------
  nuisance_vars <- c(
    "asset_precip_3m_scaled",
    "asset_precip_annual_scaled",
    "asset_precip_annual_scaled_lagged",
    "valley_precip_3m_scaled",
    "valley_precip_annual_scaled",
    "valley_precip_annual_scaled_lagged",
    "basin_precip_3m_scaled",
    "basin_precip_annual_scaled",
    "basin_precip_annual_scaled_lagged",
    "swc_australia_scaled",
    "swc_australia_scaled_lagged"
  )

  scenarios <- valley_climate_projections_with_wetlands |>
    filter(Wetland == wetland) |>
    arrange(Scenario) |>
    select(Scenario, pct_change) |>
    left_join(
      basin_flow_mean_sd |>
        select(Scenario, basin_flow_zscore_3m, basin_flow_zscore_annual),
      by = "Scenario"
    ) |>
    mutate(
      wetland_flow_3m_scaled = (scaling$wetland_mean_flow_3m * pct_change) /
        scaling$wetland_sd_flow_3m,
      wetland_flow_annual_scaled = (scaling$wetland_mean_flow_annual *
        pct_change) /
        scaling$wetland_sd_flow_annual,
      wetland_flow_annual_scaled_lagged = wetland_flow_annual_scaled,
      total_basin_flow_3m_scaled = basin_flow_zscore_3m,
      total_basin_flow_annual_scaled = basin_flow_zscore_annual,
      total_basin_flow_annual_scaled_lagged = basin_flow_zscore_annual,
      Wetland = wetland
    ) |>
    select(-basin_flow_zscore_3m, -basin_flow_zscore_annual)

  pred_newdata <- bind_rows(
    tibble(
      Scenario = "Baseline",
      pct_change = NA_real_,
      wetland_flow_3m_scaled = 0,
      wetland_flow_annual_scaled = 0,
      wetland_flow_annual_scaled_lagged = 0,
      total_basin_flow_3m_scaled = 0,
      total_basin_flow_annual_scaled = 0,
      total_basin_flow_annual_scaled_lagged = 0,
      Wetland = wetland
    ),
    scenarios
  )

  pred_newdata[nuisance_vars] <- 0

  if (!is.null(spp_code)) {
    pred_newdata$spp_code <- spp_code
  }
  if (!is.null(fx_group)) {
    pred_newdata$fx_group <- fx_group
  }

  # --- Posterior draws ------------------------------------------------------
  draws <- posterior_epred(model, newdata = pred_newdata, re_formula = NULL) |>
    t() |>
    as_tibble() |>
    mutate(scenario = pred_newdata$Scenario) |>
    pivot_longer(-scenario, names_to = ".draw", values_to = ".value")

  # --- Percent change (shared computation) ----------------------------------
  draws_with_pct <- draws |>
    group_by(.draw) |>
    mutate(
      baseline_val = .value[scenario == "Baseline"],
      percent_change = (.value - baseline_val) / baseline_val * 100
    ) |>
    ungroup()

  

  summary_results <- draws_with_pct |>
    group_by(scenario) |>
    median_hdci(percent_change, .value) |>
    mutate(wetland_name = wetland)
  if (!is.null(spp_code)) {
    summary_results$spp_code <- spp_code
  }
  if (!is.null(fx_group)) {
    summary_results$fx_group <- fx_group
  }
  list(
    summary = summary_results
  )
}
```

# Libraries and data 

```{r}
#| message: false
#| warning: false
#| label: "libraries"
#| code-fold: true
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(eawsAnalysis)
library(sf)
library(brms)
library(tidybayes)
library(patchwork)
```



```{r}
#| message: false
#| warning: false
#| label: "Load data"
#| code-fold: true
valley_climate_projections_with_wetlands <- read.csv("data/valley_climate_projections_with_wetlands.csv") |> 
    na.omit() # Scalars for each wetland, based on the valley it is located in for each climate scenario

combined_flow_metadata <- read.csv("data/combined_flow_metadata.csv") # This is the mean and sd for every wetland, based on the scaling we did for the models

basin_flow_mean_sd <- read.csv("data/basin_flow_mean_sd.csv") # This is the mean and sd for the basin flow, which we use for scaling the basin flow variables in the climate data
data(assets)
assets_coords <- assets %>%
  # Ensure we have the center point of each polygon
  mutate(centroid_lat = st_coordinates(st_centroid(geom))[, 2]) %>%
  st_drop_geometry() %>%
  # Sort by latitude descending (North to South)
  arrange(desc(centroid_lat)) %>%
  pull(Wetland) %>%
  str_to_lower()

```

# Walk through for one wetland

We get the first wetland, and load in the model
```{r}
#| message: false
#| warning: false
#| label: "load_model"
wetlands <- valley_climate_projections_with_wetlands |> 
  pull(Wetland) %>%
  unique() 
wetland <- wetlands[1]
overall_abund <- readr::read_rds("data/m1.1.rds")
print(paste0("Wetland: ", wetland))
```

Get the scaling factors for that wetland (mean and sd), because we scaled the input variables on a wetland basis. 

```{r}
#| message: false
#| warning: false
#| label: "scaling"
scaling <- combined_flow_metadata |>
    filter(wetlands == wetland) |>
    slice(1) |>
    select(
      wetland_mean_flow_3m,
      wetland_sd_flow_3m,
      wetland_mean_flow_annual,
      wetland_sd_flow_annual
    )
glimpse(scaling)
```

Scale the predictors by the appropriate climate scalars, and then convert to z-scores using the wetland-specific mean and sd. We also add in the basin flow z-scores, which are used as predictors in the model, and are scaled using the mean and sd from the `basin_flow_mean_sd` dataset.

```{r}
#| message: false
#| warning: false
#| label: "scenarios"
#| code-fold: true

scenarios <- valley_climate_projections_with_wetlands |>
    filter(Wetland == wetland) |>
    arrange(Scenario) |>
    select(Scenario, pct_change) |>
    left_join(
      basin_flow_mean_sd |>
        select(Scenario, basin_flow_zscore_3m, basin_flow_zscore_annual),
      by = "Scenario"
    ) |>
    mutate(
      wetland_flow_3m_scaled = (scaling$wetland_mean_flow_3m * pct_change) /
        scaling$wetland_sd_flow_3m,
      wetland_flow_annual_scaled = (scaling$wetland_mean_flow_annual *
        pct_change) /
        scaling$wetland_sd_flow_annual,
      wetland_flow_annual_scaled_lagged = wetland_flow_annual_scaled,
      total_basin_flow_3m_scaled = basin_flow_zscore_3m,
      total_basin_flow_annual_scaled = basin_flow_zscore_annual,
      total_basin_flow_annual_scaled_lagged = basin_flow_zscore_annual,
      Wetland = wetland
    ) |>
    select(-basin_flow_zscore_3m, -basin_flow_zscore_annual)


```


Build the new data for predicting on. We set all the target flow variables for the baseline scenario to 0 (the mean), and then add in the scaled flow variables for the scenarios. We also set all the other variables to 0 (mean).

```{r}
#| message: false
#| warning: false
#| label: "predictors"
#| code-fold: true
nuisance_vars <- c(
    "asset_precip_3m_scaled",
    "asset_precip_annual_scaled",
    "asset_precip_annual_scaled_lagged",
    "valley_precip_3m_scaled",
    "valley_precip_annual_scaled",
    "valley_precip_annual_scaled_lagged",
    "basin_precip_3m_scaled",
    "basin_precip_annual_scaled",
    "basin_precip_annual_scaled_lagged",
    "swc_australia_scaled",
    "swc_australia_scaled_lagged"
  )

pred_newdata <- bind_rows(
    tibble(
        Scenario = "Baseline",
        pct_change = NA_real_,
        wetland_flow_3m_scaled = 0,
        wetland_flow_annual_scaled = 0,
        wetland_flow_annual_scaled_lagged = 0,
        total_basin_flow_3m_scaled = 0,
        total_basin_flow_annual_scaled = 0,
        total_basin_flow_annual_scaled_lagged = 0,
        Wetland = wetland
    ),
    scenarios
)

pred_newdata[nuisance_vars] <- 0
glimpse(pred_newdata)

```

Now we predict. We use `posterior_epred()`, not `posterior_predict()` to eliminate noise from residuals, and just focus on the estimate for the mean.  
We also calculate the estimated percent change from the baseline.

```{r}
#| message: false
#| warning: false
#| label: "predictions"
model <- overall_abund



draws <- posterior_epred(model, newdata = pred_newdata, re_formula = NULL) |>
    t() |>
    as_tibble() |>
    mutate(scenario = pred_newdata$Scenario) |>
    pivot_longer(-scenario, names_to = ".draw", values_to = ".value")

draws_with_pct <- draws |>
    group_by(.draw) |>
    mutate(
      baseline_val = .value[scenario == "Baseline"],
      percent_change = (.value - baseline_val) / baseline_val * 100
    ) |>
    ungroup()

  

summary_results <- draws_with_pct |>
    group_by(scenario) |>
    median_hdci(percent_change, .value, na.rm = TRUE) |>
    mutate(wetland_name = wetland)
summary_results

DT::datatable(
  summary_results,
  options = list(
    pageLength = 7,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='70%') |> 
     DT::formatRound(columns = c("percent_change", "percent_change.lower", "percent_change.upper", ".value", ".value.lower", ".value.upper"), digits = 2) 
```

Make a plot to show the results (up to you how you want to visualise)
```{r}
#| message: false
#| warning: false
#| label: "plot"
p1 <- ggplot(summary_results |> filter(scenario != "Baseline") |> mutate(scenario = factor(scenario, levels = c("S1", "S4", "S2", "S5", "S3", "S6"))), aes(x = scenario, y = percent_change)) +
    geom_point() +
    geom_errorbar(aes(ymin = percent_change.lower, ymax = percent_change.upper), width = 0.2) +
    labs(
        x = "Climate Scenario",
        y = "Percent Change from Baseline"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(summary_results  |> mutate(scenario = factor(scenario, levels = c("Baseline", "S1", "S4", "S2", "S5", "S3", "S6"))), aes(x = scenario, y = .value)) +
    geom_point() +
    geom_errorbar(aes(ymin = .value.lower, ymax = .value.upper), width = 0.2) +
    labs(
        x = "Climate Scenario",
        y = "Absolute Change"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
p1+p2
```


# All wetlands
We wrap this all in a function - `get_climate_predictions()` and loop through the wetlands

```{r}
#| message: false
#| warning: false
#| label: "all_wetlands"

wetlands <- valley_climate_projections_with_wetlands |> 
  pull(Wetland) %>%
  unique() 

overall_abund <- readRDS("data/m1.1.rds")
results <- vector("list", length(wetlands))

for (i in 1:length(wetlands)) {
  # print(wetlands[i])
  
  results[[i]] <- get_climate_predictions(
    wetland = wetlands[i],
    model = overall_abund
  )
}

combined_df_overall <- purrr::map(results, "summary") %>% 
  bind_rows()
```

Plots like this are in the report, but you can make your own visualisation if you want.

```{r}
#| message: false
#| warning: false
#| label: "plot_all_wetlands"
#| fig-width: 20
#| fig-height: 12
ggplot2::ggplot() +
    ggplot2::geom_pointrange(
  data = combined_df_overall |> filter(scenario != "Baseline") |> mutate(scenario = factor(scenario, levels = c("S1", "S4", "S2", "S5", "S3", "S6")), wetland_name = factor(wetland_name, levels = rev(assets_coords))),
  ggplot2::aes(
  x = percent_change,
  xmin = percent_change.lower,
  xmax = percent_change.upper,
  y = wetland_name
  ),
  # shape = 17,
  # size = 7
  ) +
    facet_wrap(~scenario, nrow = 1, scales = "free_x",labeller = ggplot2::label_wrap_gen(width = 18)) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  ggplot2::geom_hline(yintercept = 22.5, color = "black", size = 0.5, linetype = "dashed", alpha = 0.8) +
  ggplot2::scale_y_discrete(
    labels = str_to_title
  ) +
  my_theme() +
  ggplot2::labs(
  x = "Percent change",
  y = "Asset"
  ) +
  ggplot2::theme(
  axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
  panel.spacing = ggplot2::unit(2, "lines"),
  panel.background = ggplot2::element_rect(fill = "#F2F2F2"),
  plot.subtitle = ggplot2::element_text(size = 14),
  axis.text.y = ggtext::element_markdown(size = 24),
  legend.position = "none"
  ) 

```

