---
title: "Task 4 Sample analysis"
author: "Will Wright"
format: 
  html:
    code-fold: false
    theme: cosmos
---


# Libraries

```{r}
#| message: false
#| warning: false
#| label: "libraries"
#| code-fold: true
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(eawsAnalysis)
library(sf)
library(brms)
library(tidybayes)
library(patchwork)
wetlands_with_valleys <- read.csv("data/wetlands_with_valleys.csv") 

```


# Data
Again, have omitted cleaning information here. See Github repo for full details.
We need to pull in all the other data because we want to keep all variables at the levels they were in a given year, and only manipulate the flow variables.
```{r}
#| message: false
#| warning: false
#| label: "data"
#| code-fold: true
# Rainfall data
asset_precip <- read.csv("data/asset_precip_data.csv")
valley_precip <- read.csv("data/valley_precip_data.csv")
precip <- read.csv("data/precip_data.csv")

# SWC (flooding) data
swc <- read.csv("data/australia_annual_water_area_presence.csv")
wetlands_swc <- read.csv("data/wetland_inundation_results.csv")
basin_swc <- read.csv("data/basin_inundation_results.csv") 
valleys_swc <- read.csv("data/valley_inundation_results.csv")

# Flow data
combined_flow_data <- read.csv("data/combined_flow_data.csv")
combined_flow_metadata <- read.csv("data/combined_flow_metadata.csv")
total_flow <- read.csv("data/total_flow.csv")

```

## Eflow data
I give a short overview of the cleaning of these data here. For other wetlands, this may not be the same, but the required outputs will be similar. 

Here's the raw data
```{r}
#| message: false
#| warning: false
#| label: "raw"
ewater <- read.csv("data/eflows.csv")
glimpse(ewater)
```

The data is grouped by month, and has a bunch of different water classifications. What we really care about is the total flow, and the non-environmental flow in the year or three months preceeding the survey.

We sum by year and get the 3 month sum for each year (across all columns). Then we create a new column for non-environmental flow, which is the total flow minus the environmental flow. So for every year, we end up with a total flow, environmental flow, and non-environmental flow for the year and the 3 months preceeding the survey.
```{r}
#| message: false
#| warning: false
#| label: "Eflow data"
#| code-fold: true
ewater_wy <- ewater |>
  mutate(
    month_num = match(month, month.name),
    date = as.Date(paste(year, month_num, 1, sep = "-")),
    # Nov-Oct water year: Nov+ belongs to following year's Oct label
    water_year_nov = if_else(month_num >= 11, year + 1L, as.integer(year))
  )

# Annual summary
ewater_annual <- ewater_wy |>
  summarise(
    across(c(marebone_total, env_flow, EWA, FMZ, irrigation, baseflow, other), sum, na.rm = TRUE),
    .by = water_year_nov
  ) |>
  arrange(water_year_nov)

ewater_annual <- ewater_annual |>
  mutate(
    non_env_flow = marebone_total - env_flow 
  ) |> 
    select(water_year_nov, non_env_flow, env_flow, marebone_total)

# Now the 3m
ewater_wy_3m <- ewater |>
  mutate(
    month_num = match(month, month.name),
    date = as.Date(paste(year, month_num, 1, sep = "-")),
    # Nov-Oct water year: Nov+ belongs to following year's Oct label
    water_year_nov = if_else(month_num >= 11, year + 1L, as.integer(year))
  ) %>% 
    filter(month_num >= 7, month_num < 10) # preceding 3 months

ewater_annual_3m <- ewater_wy_3m |>
  summarise(
    marebone_total_3m = sum(marebone_total, na.rm = TRUE),
    env_flow_3m = sum(env_flow, na.rm = TRUE),
    EWA_3m = sum(EWA, na.rm = TRUE),
    FMZ_3m = sum(FMZ, na.rm = TRUE),
    irrigation_3m = sum(irrigation, na.rm = TRUE),
    baseflow_3m = sum(baseflow, na.rm = TRUE),
    other_3m = sum(other, na.rm = TRUE),
    .by = water_year_nov
  ) |>
  arrange(water_year_nov)

ewater_annual_3m <- ewater_annual_3m |>
  mutate(
    non_env_flow_3m = marebone_total_3m - env_flow_3m
  ) |> 
    select(water_year_nov, non_env_flow_3m, env_flow_3m, marebone_total_3m)

ewater_annual <- ewater_annual |>
  left_join(ewater_annual_3m, by = "water_year_nov")
ewater_annual
```

We do some renaming to make the columns match the training data, and also grab the scaling data
```{r}
#| message: false
#| warning: false
#| label: "scaling"
#| code-fold: true
mac_scaling <- combined_flow_metadata |>
    filter(wetlands == "macquarie marshes") |>
    slice(1) |>
    select(
      wetland_mean_flow_3m,
      wetland_sd_flow_3m,
      wetland_mean_flow_annual,
      wetland_sd_flow_annual
    )
ewater_annual <- ewater_annual |>
  mutate(
    wetland_flow_annual_scaled_total   = (marebone_total - mac_scaling$wetland_mean_flow_annual) / mac_scaling$wetland_sd_flow_annual,
    wetland_flow_annual_scaled_non_env = ((non_env_flow) - mac_scaling$wetland_mean_flow_annual) / mac_scaling$wetland_sd_flow_annual,
    wetland_flow_annual_scaled_total_lagged = lag(wetland_flow_annual_scaled_total),
    wetland_flow_annual_scaled_non_env_lagged = lag(wetland_flow_annual_scaled_non_env),
    wetland_flow_3m_scaled_total   = (marebone_total_3m - mac_scaling$wetland_mean_flow_3m) / mac_scaling$wetland_sd_flow_3m,
    wetland_flow_3m_scaled_non_env = ((non_env_flow_3m ) - mac_scaling$wetland_mean_flow_3m) / mac_scaling$wetland_sd_flow_3m,
    wetland_flow_3m_scaled_total_lagged = lag(wetland_flow_3m_scaled_total),
    wetland_flow_3m_scaled_non_env_lagged = lag(wetland_flow_3m_scaled_non_env)
    
  )


```

## Other variables
We load in other variables that we need
```{r}
#| message: false
#| warning: false
#| code-fold: true
#| label: "other variables"

full_data <- tibble(surv_year = seq(2010, 2024), Wetland = "macquarie marshes")  %>%
  left_join(
    asset_precip %>% mutate(Wetland = str_to_lower(Wetland)),
    join_by(surv_year == year, Wetland == Wetland)
  ) %>%
  left_join(wetlands_with_valleys) %>% 
  mutate(ValleyName = str_to_lower(ValleyName)) %>%
  left_join(
    valley_precip,
    join_by(surv_year == year, ValleyName == valleyname)
  ) %>%
  left_join(precip, by = join_by("surv_year" == "year")) %>%
  left_join(
    combined_flow_data,
    by = join_by("surv_year" == "water_year", Wetland == wetlands)
  ) %>%
  filter(
    !is.na(wetland_flow_annual_scaled)
  ) %>%
  left_join(total_flow, by = join_by("surv_year" == "water_year")) %>%
  left_join(swc, join_by('surv_year' == "year")) %>%
  select(
    -c(
      subwetland,
      asset_precip_3m,
      asset_precip,
      X,
      valley_precip_3m,
      valley_precip,
      basin_precip_3m,
      basin_precip,
      flow_3m,
      flow,
      water_area_m2
    )
  ) %>%
  mutate(id = row_number())
```

## Make new counterfactual data
and make the coutnerfactual datasets - one with environmental flow, one without
```{r}
#| message: false
#| warning: false
#| label: "new data"
#| code-fold: true
new_data_with_env <- full_data %>% 
  filter(Wetland == "macquarie marshes") %>% 
  select(
    -c(
      wetland_flow_3m_scaled,
      wetland_flow_annual_scaled,
      wetland_flow_annual_scaled_lagged
    )
  ) %>%
  left_join(ewater_annual, by = join_by("surv_year" == "water_year_nov")) %>% 
  rename(
    wetland_flow_3m_scaled = wetland_flow_3m_scaled_total,
    wetland_flow_annual_scaled = wetland_flow_annual_scaled_total,
    wetland_flow_annual_scaled_lagged = wetland_flow_annual_scaled_total_lagged
  ) %>% 
    select(
        -c(
          marebone_total,
          env_flow,
          non_env_flow,
          marebone_total_3m,
          env_flow_3m,
          
          wetland_flow_3m_scaled_non_env,
          wetland_flow_3m_scaled_non_env_lagged,
          wetland_flow_annual_scaled_non_env,
          wetland_flow_annual_scaled_non_env_lagged
        )
    )|> mutate(across(everything(), ~replace_na(.x, 0)))|> ungroup()


new_data_no_env <- full_data |> 
 
  filter(Wetland == "macquarie marshes") %>% 
  select(
    -c(
      wetland_flow_3m_scaled,
      wetland_flow_annual_scaled,
      wetland_flow_annual_scaled_lagged
    )
  ) %>%
  left_join(ewater_annual, by = join_by("surv_year" == "water_year_nov")) %>% 
  rename(
    wetland_flow_3m_scaled = wetland_flow_3m_scaled_non_env,
    wetland_flow_annual_scaled = wetland_flow_annual_scaled_non_env,
    wetland_flow_annual_scaled_lagged = wetland_flow_annual_scaled_non_env_lagged
  ) %>% 
    select(
        -c(
          marebone_total,
          env_flow,
          
          non_env_flow,
          marebone_total_3m,
          env_flow_3m,
          
          non_env_flow_3m,
          
          wetland_flow_3m_scaled_total,
          wetland_flow_3m_scaled_total_lagged,
          
          wetland_flow_annual_scaled_total,
          wetland_flow_annual_scaled_total_lagged

        )
    ) |> mutate(across(everything(), ~replace_na(.x, 0)))|> ungroup()

```


# Modelling

Now we can run the models and make some plots/tables
```{r}
#| message: false
#| warning: false
#| label: "models"
#| code-fold: true
overall_abund <- readRDS("data/m1.1.rds")

predicted_with_env <- overall_abund %>%
  epred_draws(newdata = new_data_with_env, re_formula = NULL)
predicted_no_env <- overall_abund %>%
  epred_draws(newdata = new_data_no_env, re_formula = NULL)

output_df <- predicted_with_env |>
  ungroup() |>
  filter(surv_year != 2024) |>
  select(surv_year, .draw, .epred) |>
  inner_join(
    predicted_no_env |>
      filter(surv_year != 2024) |>
      select(surv_year, .draw, .epred_no_env = .epred),
    by = join_by(surv_year, .draw)
  ) |>
  mutate(
    diff = .epred - .epred_no_env,
    pct_diff = (.epred - .epred_no_env) / .epred * 100
  ) |>
  group_by(surv_year) |>
  median_hdci(diff, pct_diff, .width = 0.95)

overall_abund_pct <- output_df |> 
  ggplot(aes(x = surv_year, y = pct_diff )) +
  geom_ribbon(aes(ymin = pct_diff.lower, ymax = pct_diff.upper), fill = "lightblue", alpha = 0.5) +
  geom_line(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Survey Year",
    y = "Percentage Difference"
  ) +
  my_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


Here's the table
```{r}
#| message: false
#| warning: false
#| echo: false
#| label: "table"
DT::datatable(
  output_df,
  options = list(
    pageLength = 15,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='70%') |> 
     DT::formatRound(columns = c("pct_diff", "pct_diff.lower", "pct_diff.upper", "diff", "diff.lower", "diff.upper"), digits = 2) 

```

and the plot
```{r}
#| message: false
#| warning: false
#| echo: false
#| label: "plot"
overall_abund_pct

```



