---
title: "Waterbird Analysis Workshop"
author: "Will Wright"
date: today
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    code-fold: true
    fig-width: 10
    fig-height: 6
---

## Workshop Overview

This workshop covers waterbird analysis in the Murray-Darling Basin:

1. **Trend Analysis** — Extract data and identify population trends
2. **Model Building** — Fit Bayesian hurdle-gamma models
3. **Climate Predictions** — Project waterbird responses under climate scenarios
4. **E-flow Analysis** — Macquarie Marshes environmental flow assessment

# Trend Analysis {background-color="#000000"}

## The `eawsAnalysis` Package

```{r}
#| echo: true
#| eval: false
library(eawsAnalysis)
data(data_clean)
```

Purpose-built package for Eastern Australian Waterbird Survey analysis:

- Data cleaning and filtering
- Mann-Kendall trend tests
- Visualization tools
- Integrated workflow functions

## Running Trend Analysis

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "2-8"
result <- clean_data_run_models(
  dataset = data_clean,
  programs = c("Eastern Australian Survey"),
  metric = "abundance",
  wetlands = NULL,
  valleys = NULL,
  grouping_cols = "spp_code",
  decade_analysis = FALSE
)
```

Returns cleaned data, plots, and summary statistics

## Trend Results

```{r}
#| echo: false
#| eval: false
plot <- result[[2]]
plot
```

- Species-level trends over time
- Mann-Kendall tau and p-values
- Long-term averages and variability

# Model Building {background-color="#000000"}

## Model Structure

**Hurdle-Gamma Model** for zero-inflated count data:

::: {.incremental}
- **Hurdle component** — Binomial model for presence/absence
- **Gamma component** — Continuous positive values when present
- **Bayesian estimation** — Full uncertainty quantification via MCMC
:::

## Environmental Predictors

::: {.columns}
::: {.column width="50%"}
**Flow**
- Wetland-level annual flow
- Basin-level flow (3-month)
- Lagged effects

**Precipitation**
- Asset-level rainfall
- Valley-level rainfall
- Annual totals
:::

::: {.column width="50%"}
**Surface Water**
- Wetland inundation
- Basin inundation
- Valley inundation
- Lagged conditions
:::
:::

## Prior Specification

```{r}
#| echo: true
#| eval: false
priors <- c(
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sigma")
)
```

Weakly informative priors allow data to dominate

## Model Fitting

```{r}
#| echo: true
#| eval: false
#| output: false
m1.1 <- brm(
  full_formula,
  family = hurdle_gamma(),
  data = full_data,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 4000
)
```

## Model Diagnostics

Check convergence and fit:

- **Trace plots** — MCMC chain mixing
- **R-hat values** — Should be < 1.01
- **Posterior predictive checks** — Compare simulated to observed data
- **Bayesian R²** — Variance explained

# Climate Predictions {background-color="#000000"}

## Climate Scenarios

Six projection scenarios from valley-level climate models:

- **Historical baseline**
- **SSP2-4.5** (moderate emissions)
- **SSP3-7.0** (high emissions)
- Multiple GCM projections

Precipitation changes translated to flow z-scores

## Prediction Workflow

```{r}
#| echo: true
#| eval: false
scenarios <- get_climate_predictions(
  model = m1.1,
  wetland = "macquarie marshes",
  newdata = pred_newdata,
  metric = "abundance"
)
```

Generate posterior predictions for each scenario

## Expected Changes

```{r}
#| echo: false
#| eval: false
# Plot showing percent change across scenarios
```

Quantify expected waterbird response:

- Median change from baseline
- Credible intervals
- Probability of decline/increase

# E-flow Analysis {background-color="#000000"}

## Macquarie Marshes

Environmental flow assessment:

- Simulate different flow scenarios
- Predict waterbird outcomes
- Identify optimal flow regimes
- Support adaptive management

## Take-Home Messages

::: {.incremental}
1. **Integrated workflow** — `eawsAnalysis` package streamlines analysis
2. **Bayesian models** — Full uncertainty quantification for predictions
3. **Climate-ready** — Framework for scenario planning
4. **Management-relevant** — E-flow tools support decision-making
:::

## Questions?

**Resources:**

- GitHub: [github.com/USERNAME/waterbirds-workshop](https://github.com/USERNAME/waterbirds-workshop)
- Package: `remotes::install_github("will02-code/eawsAnalysis")`
- Documentation: [Link to your website]

---

Thank you!
