---
title: "Trend analysis with `eawsAnalysis`"
---



```{r}
# install.packages("remotes")
# remotes::install_github("will02-code/eawsAnalysis")
library(eawsAnalysis)
library(magrittr)
library(readr)


```



#### Example analysis

We'll walk through an example analysis using the `eawsAnalysis` package. The same process was repeated for all metrics and scales.

There are four main steps to the analysis: 1. Define parameters 2. Clean data 3. Run models 4. Clean output

We have provided code snippets for each step below, but generally we recommend using the `clean_data_run_models()` function which combines all steps into one function call.

##### Define parameters

Here we define the parameters for the analysis.

```{r}
#| label: Define parameters
#| code-summary: "Define analysis parameters"
#| eval: false
programs = c("Eastern Australian Survey") # <1>
dataset = data_clean                      # <2>
metric = "abundance"                      # <3>
wetlands = NULL                           # <4>
valleys = NULL                            # <5>
basinDiv = NULL                           # <5>
grouping_cols = "spp_code"                # <6>
decade_analysis = FALSE                   # <7>
by_condition = NULL                       # <8>
```
1. Survey programs to include.
2. Dataset to use.
3. Metric to analyse.
4. Wetlands to include (`NULL` for all).
5. Valleys and basin divisions to include (`NULL` for all).
6. Grouping column (`NULL` for overall analysis).
7. Whether to run decade-wise analysis (`TRUE`/`FALSE`).
8. Surface water condition column to group by (e.g. `"basin_condition"`, `"wetland_condition"`); `NULL` for no grouping.

##### Clean data

We need the data to be cleaned before running models. This involves filtering to the specified programs, wetlands, valleys, and grouping column, as well as calculating the metric of interest. See `get_bird_data()` function for details.

```{r}
#| label: Clean data
#| code-summary: "Clean data for analysis"
#| eval: false

d2 <- get_bird_data(
  dataset,
  programs = programs,
  metric = metric,
  wetlands = wetlands,
  valleys = valleys,
  basinDiv = basinDiv,
  grouping_cols = grouping_cols,
  by_condition = by_condition

)

```

##### Run models

We then run the specified models on the cleaned data. See `run_mk_models()` function for details. We use the grouping column, and decade analysis parameters.

```{r}
#| label: Run models
#| code-summary: "Run statistical models"
#| eval: false
output <- run_mk_models(
  d2,

  decade_analysis = decade_analysis,
  grouping_cols = grouping_cols,
  
)


```


##### Clean output

Finally, we clean the output from the models to extract relevant results, trends, and predictions. See `clean_model_output()` function for details.

```{r}
#| label: Clean output
#| code-summary: "Clean model output"
#| eval: false

cleaned <- clean_model_output(
  output,
  decade_analysis = decade_analysis,
  grouping_cols = grouping_cols
)


```

##### Make plots and summary stats

We can then make plots of the fitted models and calculate summary statistics using the `post_fitting_line_plot()` function and summarisation code within `clean_data_run_models()`.

```{r}
#| label: Plots and summary stats
#| code-summary: "Generate plots and summary statistics"
#| eval: false

group_vars <- c(
    if (decade_analysis) "decade",
    if (!is.null(grouping_cols)) grouping_cols,
    if (!is.null(by_condition)) by_condition
  )

cleaned_summary <- cleaned %>%
    dplyr::ungroup() %>%
    dplyr::summarise(
      min = min(value),
      q1 = stats::quantile(value, 0.25),
      median = stats::median(value),
      q3 = stats::quantile(value, 0.75),
      max = max(value),
      long_term_avg = mean(value),
      dplyr::across(
        tidyselect::any_of(c(
          "percent_estimate",
          "percent_p.value",
          "mk_tau",
          "mk_p_value",

          'mk_logslope',
          'mk_percentslope',
          'mk_intercept',
          "n_rows"
        )),
        dplyr::first
      ),

      .by = tidyselect::all_of(group_vars)
    )

  p <- post_fitting_line_plot(
    cleaned,
    decade_analysis = decade_analysis,
    grouping_cols = grouping_cols,
    metric = metric
  )
  p
```

###### Complete analysis function

To simplify the analysis process, there is a wrapper function `clean_data_run_models()` which combines all steps into one function call. This function takes in the dataset and analysis parameters, cleans the data, runs the models, cleans the output, generates plots, and calculates summary statistics.

```{r}
#| label: Complete analysis function
#| code-summary: "Run complete analysis in one function"
result <- clean_data_run_models(
  dataset = data_clean,
  programs = c("Eastern Australian Survey"),
  metric = "abundance",
  wetlands = NULL,
  valleys = NULL,
  grouping_cols = NULL, # Grouping column (e.g. "spp_code" for species-level analysis)
  basinDiv = NULL, 
  decade_analysis = FALSE, 
  trend_line = TRUE # Whether to generate trend line plot


)
cleaned_data <- result[[1]]
plot <- result[[2]]
summary_stats <- result[[3]]
```

So we end up with three outputs: - `cleaned_data`: Cleaned data frame with model results and predictions - `plot`: ggplot2 object of fitted model plot - `summary_stats`: Data frame containing compiled trend statistics (e.g., Mann-Kendall tau, p-values).

**Full data frame**


```{r}
#| label: print_cleaned_data
#| code-summary: "Print cleaned data"
#| echo: false

DT::datatable(
  cleaned_data,
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='60%') |> 
     DT::formatRound(columns = c("mk_tau", "mk_logslope", "mk_percentslope", "mk_intercept"), digits = 2) |> 
     DT::formatRound(columns = c("mk_p_value"), digits = 3)
```

**Fitted model plot**

```{r}
#| label: fitted_model_plot
#| echo: false
plot
```

**Summary statistics**

```{r}
#| label: print_summary_stats
#| code-summary: "Print summary statistics"
#| echo: false



DT::datatable(
  summary_stats,
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='60%')|> 
     DT::formatRound(columns = c("mk_tau", "mk_logslope", "mk_percentslope", "mk_intercept", "long_term_avg"), digits = 2) |> 
     DT::formatRound(columns = c("mk_p_value"), digits = 3)
```


##### Or with functional groups


```{r}
#| label: complete-analysis-fx-group
#| code-summary: "Run complete analysis in one function"
result_fx <- clean_data_run_models(
  dataset = data_clean,
  programs = c("Eastern Australian Survey"),
  metric = "abundance",
  wetlands = NULL,
  valleys = NULL,
  grouping_cols = "fx_group",
  basinDiv = NULL, 
  decade_analysis = FALSE, 
  trend_line = TRUE
)
cleaned_data_fx <- result_fx[[1]]
plot_fx <- result_fx[[2]]
summary_stats_fx <- result_fx[[3]]
```

**Full data frame**

```{r}
#| label: print-cleaned-data-fx-group
#| code-summary: "Print cleaned data"
#| echo: false

DT::datatable(
  cleaned_data_fx,
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='60%') |> 
      
     DT::formatRound(columns = c("mk_tau", "mk_logslope", "mk_percentslope", "mk_intercept"), digits = 2) |> 
     DT::formatRound(columns = c("mk_p_value"), digits = 3)
```

**Fitted model plot**

```{r}
#| label: fitted-model-plot-fx-group
#| fig-cap: "Fitted Model Plot"
#| fig-cap-location: top
#| echo: false
plot_fx
```

**Summary statistics**

```{r}
#| label: print-summary-stats-fx-group
#| code-summary: "Print summary statistics"
#| echo: false
DT::datatable(
  summary_stats_fx,
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='60%') |> 
     DT::formatRound(columns = c("mk_tau", "mk_logslope", "mk_percentslope", "mk_intercept", "long_term_avg"), digits = 2) |> 
     DT::formatRound(columns = c("mk_p_value"), digits = 3)


```

##### We can also get trends by SWC 
To do this, we need to calculate SWC first and join it to the dataset

```{r}
#| label: calculate-swc
#| code-summary: "Calculate surface water condition (SWC) and join to dataset"
#| echo: false
#| code-fold: true
classify_swc <- function(df, sw_col = "surface_water", lag_col = "surface_water_lagged", 
                          group_col = NULL, output_col = "condition") {
  df |>
    dplyr::group_by(across(all_of(group_col))) |>
    dplyr::mutate(
      sw_p  = dplyr::percent_rank(.data[[sw_col]])  * 100,
      ant_p = dplyr::percent_rank(.data[[lag_col]]) * 100
    ) |>
    dplyr::ungroup() |>
    dplyr::filter(!is.na(ant_p)) |>
    dplyr::mutate(
      "{output_col}" := dplyr::case_when(
        sw_p <= 15 & ant_p <= 15  ~ "very dry",
        sw_p <= 15 & ant_p <= 45  ~ "very dry",
        sw_p <= 15 & ant_p <= 60  ~ "dry",
        sw_p <= 15 & ant_p <= 85  ~ "dry",
        sw_p <= 15 & ant_p >  85  ~ "moderate",
        sw_p <= 45 & ant_p <= 15  ~ "very dry",
        sw_p <= 45 & ant_p <= 60  ~ "dry",
        sw_p <= 45 & ant_p <= 85  ~ "moderate",
        sw_p <= 45 & ant_p >  85  ~ "wet",
        sw_p <= 60 & ant_p <= 45  ~ "dry",
        sw_p <= 60 & ant_p <= 60  ~ "moderate",
        sw_p <= 60 & ant_p >  60  ~ "wet",
        sw_p <= 85 & ant_p <= 15  ~ "dry",
        sw_p <= 85 & ant_p <= 45  ~ "moderate",
        sw_p <= 85 & ant_p <= 85  ~ "wet",
        sw_p <= 85 & ant_p >  85  ~ "very wet",
        sw_p <= 100 & ant_p <= 15 ~ "moderate",
        sw_p <= 100 & ant_p <= 45 ~ "moderate",
        sw_p <= 100 & ant_p <= 60 ~ "wet",
        sw_p <= 100 & ant_p >  60 ~ "very wet",
        TRUE ~ "unclassified"
      )
    ) %>% 
      dplyr::select(tidyselect::all_of(group_col), year, !!rlang::sym(output_col))
}

valleys_swc <- readr::read_csv(("data/valley_inundation_results.csv")) |>
  dplyr::group_by(valley_name) |>
  dplyr::mutate(surface_water = inundated_area, surface_water_lagged = lag(inundated_area)) |>
  classify_swc(group_col = "valley_name", output_col = "valley_condition")

wetlands_swc <- readr::read_csv(("data/wetland_inundation_results.csv")) |>
  dplyr::group_by(wetland_name) |>
  dplyr::mutate(surface_water = inundated_area, surface_water_lagged = lag(inundated_area)) |>
  classify_swc(group_col = "wetland_name", output_col = "wetland_condition") %>% 
  dplyr::mutate(wetland_name = stringr::str_to_lower(wetland_name))

basin_swc <- readr::read_csv(("data/basin_inundation_results.csv")) |>
  dplyr::mutate(surface_water = inundated_area, surface_water_lagged = lag(inundated_area)) |>
  classify_swc(output_col = "basin_condition")

clean_dataset_with_swc <- data_clean %>% 
  dplyr::left_join(valleys_swc, by = dplyr::join_by(ValleyName == valley_name, surv_year == year)) %>%
  dplyr::left_join(wetlands_swc, by = dplyr::join_by(Wetland == wetland_name, surv_year == year)) %>%
  dplyr::left_join(basin_swc, by = dplyr::join_by(surv_year == year)) 




```



```{r}
#| label: run-models-with-swc
#| code-summary: "Run models with surface water condition (SWC)"
#| echo: false

result_fx_swc <- clean_data_run_models(
  dataset = clean_dataset_with_swc,
  programs = c("Eastern Australian Survey"),
  metric = "abundance",
  wetlands = NULL,
  valleys = NULL,
  grouping_cols = "fx_group",
  basinDiv = NULL, 
  decade_analysis = FALSE, 
  trend_line = TRUE,
  by_condition = "basin_condition"
)
DT::datatable(
  result_fx_swc[[3]],
  options = list(
    pageLength = 25,
    autoWidth = TRUE,
    scrollX = TRUE,
    dom = 'lrtip'  # Remove the filter box if you want more compact layout
  ),
  class = 'compact stripe'
) |> 
     DT::formatStyle( 0, target= 'row',lineHeight='60%')|> 
     DT::formatRound(columns = c("mk_tau", "mk_logslope", "mk_percentslope", "mk_intercept", "long_term_avg"), digits = 2) |> 
     DT::formatRound(columns = c("mk_p_value"), digits = 3)

```

